---
title: "Differential Expression Analysis of HSC and progenitor cells"
author: "Paula Vela Moreno"
output: html_document
date: "2025-05-05"
---

#############################################################
#                                                           #
#   Diferential expression analysis on 106 hematopoietic    #
#   stem and progenitor cells to identify stable            #
#   housekeeping genes.                                     #
#   Data: From GEO accession number GSE233478               #
#                                                           #
#############################################################

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



*Load necessary libraries*
```{r load_libraries, include=FALSE}
# Load the necessary libraries
library(DESeq2) 
library(ggplot2) 
library(tidyr)
library(dplyr)
library(pheatmap) 
library(clusterProfiler)
library(RUVSeq)
library("org.Hs.eg.db", character.only = TRUE) # Load the genome wide annotation for Human
library(ineq)
library(ggrepel)
library(VennDiagram)
library(ggVennDiagram)
library(patchwork)
library(cowplot)
```


```{r set_directory, include = FALSE}
# Set the working directory
setwd("../Documents/BDBI/INTERNSHIP/")
```

```{r}
sessionInfo()
```


```{r load_the_data, eval=FALSE, warning=FALSE}
# Load the data
counts_106 <- read.csv("count_matrix106.csv", header = TRUE, row.names = 1, check.names = FALSE) 
metadata <- read.csv("metadata_106.csv", header=TRUE, row.names = 1)

# Rename the categories of "Donor ID" to just have the donor and not the ID
metadata$Category <- as.character(metadata$Donor.ID)
metadata$Category[metadata$Donor.ID %in% c("Fetus #11", "Fetus #10", "Fetus #12")] <- "Fetus"
metadata$Category[metadata$Donor.ID %in% c("Newborn #04", "Newborn #05", "Newborn #06", "Newborn #11")] <- "Newborn"
metadata$Category[metadata$Donor.ID %in% c("Adult #01", "Adult #02", "Adult #04")] <- "Adult"
metadata$Category[metadata$Donor.ID %in% c("Embryo #06", "Embryo #10", "Embryo #11", "Embryo #12", "Embryo #13", "Embryo #14")] <- "Embryo"

# convert columns Tissue, Cell.type, Donor.ID to factors
metadata$Tissue <- as.factor(metadata$Tissue)
metadata$Cell.type <- as.factor(metadata$Cell.type)
metadata$Donor.ID <- as.factor(metadata$Donor.ID)
metadata$Category <- factor(metadata$Category, levels = c("Fetus", "Newborn", "Adult", "Embryo"))

View(counts_106)
View(metadata)
```

```{r reordering_columns}
# Reorder the columns of counts_106 to have the same order as the rows in metadata
counts_106 <- counts_106[, rownames(metadata)]
```


```{r}
table(metadata$Donor.ID)
table(metadata$Category, metadata$Tissue)
table(metadata$Donor.ID, metadata$Tissue)
table(metadata$Cell.type, metadata$Category)
```


## 1. Construct the DESeqDataSet() object (just for initial visualization and calculation of statistics)
```{r}
dds_106 <- DESeqDataSetFromMatrix(countData = counts_106,
                                  colData = metadata,
                                  design = ~ 1) 
```



## 2. Transformation and initial PCA 
**By tissue**
```{r}
vsd_pca_106 <- vst(dds_106, blind = TRUE) # 'blind = TRUE' >> without taking into account the design 

pcaData_106 <- plotPCA(vsd_pca_106, intgroup = "Tissue", returnData = TRUE)
percentage_variation_106 <- round(100 * attr(pcaData_106, "percentVar"))

p1 <- ggplot(pcaData_106, aes(PC1, PC2, color = Tissue)) +
  geom_point(size = 3) +
  labs(title = "Initial PCA", 
       x = paste0("PC1: ", percentage_variation_106[1], "%"),
       y = paste0("PC2: ", percentage_variation_106[2], "%")) +
  scale_color_manual(values = c("#3011F7", "#16ADEE", "#A23DED", "#348E4B", "#C7E825", "#F37A1B")) +
  theme_minimal()

ggsave("Initial_PCA.png", width = 10, height = 8)
```

**By tissue (with names)**
```{r}
pcaData_106$Sample <- rownames(pcaData_106)

ggplot(pcaData_106, aes(PC1, PC2, color = Tissue)) +
  geom_point(size = 3) +
  geom_text_repel(aes(label = Sample), size = 3, max.overlaps = 20) + # add the samples names
  labs(title = "Initial PCA", 
       x = paste0("PC1: ", percentage_variation_106[1], "%"),
       y = paste0("PC2: ", percentage_variation_106[2], "%")) +
  scale_color_manual(values = c("#3011F7", "#16ADEE", "#A23DED", "#348E4B", "#C7E825", "#F37A1B")) +
  theme_minimal()

ggsave("Initial_PCA_names.png", width = 10, height = 8)
```
We found an outlier: GSM7502218
Remove it due to contamination.



## 3. Outlier removal and initial visualizations
```{r outlier_removal_GSM7502218}
# Remove the outlier from the count matrix
counts_105 <- counts_106
counts_105$GSM7502218 <- NULL # GSM7502218 is the outlier
dim(counts_105)

# Remove the outlier from the metadata
metadata_removed <- metadata
metadata_removed <- metadata_removed[rownames(metadata_removed) != "GSM7502218", ]
dim(metadata_removed)


dds_105 <- DESeqDataSetFromMatrix(countData = counts_105,
                                          colData = metadata_removed,
                                          design = ~ 1)
```

**Initial PCA without outlier**
```{r}
vsd_pca_105 <- vst(dds_105, blind = TRUE) # 'blind = TRUE' >> without taking into account the design 

pcaData_105 <- plotPCA(vsd_pca_105, intgroup = "Tissue", returnData = TRUE)
percentage_variation_105 <- round(100 * attr(pcaData_105, "percentVar"))

p1 <- ggplot(pcaData_105, aes(PC1, PC2, color = Tissue)) +
  geom_point(size = 3) +
  labs(title = "Initial PCA", 
       x = paste0("PC1: ", percentage_variation_105[1], "%"),
       y = paste0("PC2: ", percentage_variation_105[2], "%")) +
  scale_color_manual(values = c("#3011F7", "#16ADEE", "#A23DED", "#348E4B", "#CBC118", "#F37A1B")) +
  theme_minimal()

ggsave("Initial_PCA_cleaned.png", width = 10, height = 8)
```

**Tissue and developmental stage (Category)**
```{r}
vsd_pca_105 <- vst(dds_105, blind = TRUE) # 'blind = TRUE' >> without taking into account the design 

pcaData_105 <- plotPCA(vsd_pca_105, intgroup = c("Tissue","Category"), returnData = TRUE)
percentage_variation_105 <- round(100 * attr(pcaData_105, "percentVar"))

colors_tissue <- c("#CBC118", "#16ADEE", "#3011F7", "#348E4B", "#CC6A0F", "#A23DED")

p1_2 <- ggplot(pcaData_105, aes(PC1, PC2, color = Tissue, shape = Category)) +
  geom_point(size = 3, stroke = 1) +
  labs(x = paste0("PC1: ", percentage_variation_105[1], "%"),
       y = paste0("PC2: ", percentage_variation_105[2], "%"),
       shape = "Developmental stage") +
  scale_color_manual(values = colors_tissue) +
  scale_shape_manual(values = c("Embryo" = 3, "Fetus" = 2, "Newborn" = 16, "Adult" = 15)) +
  theme_minimal()

ggsave("PCA_Tissue_Category.png", width = 10, height = 8)
```


**By cell type and Tissue**
```{r}
vsd_pca_105 <- vst(dds_105, blind = TRUE) # 'blind = TRUE' >> without taking into account the design 
pcaData_105_cell_type <- plotPCA(vsd_pca_105, intgroup = c("Tissue", "Cell.type"), returnData = TRUE)
percentage_variation_105_cell_type <- round(100 * attr(pcaData_105_cell_type, "percentVar"))

colors_cellType <- c("#1b9e77", "#e66101", "#c58df2", "#f468b3", "#C7E825", "#ffcc00", "#ED7AF5", "#66a61e", "#1f78b4", "#262122", "#777777", "#a6761d")

p1_cell <- ggplot(pcaData_105_cell_type, aes(PC1, PC2, color = Cell.type, shape = Tissue)) +
  geom_point(size = 3, stroke = 1) +
  labs(x = paste0("PC1 Var: ", percentage_variation_105_cell_type[1], " % variance"),
       y = paste0("PC2 Var: ", percentage_variation_105_cell_type[2], " % variance")) +
  scale_color_manual(values = colors_cellType) +
  scale_shape_manual(values = c(16, 17, 15, 3, 2, 8)) +
  theme_minimal()

ggsave("Initial_PCA_cell-tissue.png", width = 14, height = 8)
```

**PCA embryonic tissues**
```{r}
# Tissue
vd_emb <- vsd_pca_105[, vsd_pca_105$Tissue %in% c("Aorta-gonad-mesonephros","Placenta", "Yolk Sac")]
p2 <- plotPCA(vd_emb, intgroup="Tissue") + 
  ggtitle("PCA embryonic tissues") + 
  scale_color_manual(values = c("#16ADEE", "#348E4B", "#A23DED")) + 
  theme_minimal() +
  theme(plot.title = element_text(face = "bold"))

ggsave("PCA_embryonic.png", width = 10, height = 8)



# Tissue + cell type
pca_embryo <- plotPCA(vd_emb, intgroup = c("Tissue", "Cell.type"), returnData = TRUE)
perc_emb <- round(100 * attr(pca_embryo, "percentVar"))

p2_2 <- ggplot(pca_embryo, aes(PC1, PC2, color = Tissue, shape = Cell.type)) +
  geom_point(size = 3, stroke = 1) +
  labs(title = "PCA embryonic tissues",
       x = paste0("PC1 Var: ", perc_emb[1], " % variance"),
       y = paste0("PC2 Var: ", perc_emb[2], " % variance")) +
  scale_color_manual(values = c("#16ADEE", "#348E4B", "#A23DED")) +
  scale_shape_manual(values = c(15, 3, 2)) +
  theme_minimal()

ggsave("PCA_embryonic_celltype.png", width = 10, height = 8)
```



**PCA non-embryonic tissues**
```{r}
vd_no_emb <- vsd_pca_105[, vsd_pca_105$Tissue %in% c("Adult Bone Marrow","Fetal liver", "Umbillical Cord Blood")]
p3 <- plotPCA(vd_no_emb, intgroup="Tissue") + 
  ggtitle("PCA non-embryonic tissues") + 
  scale_color_manual(values = c("#CBC118", "#3011F7", "#CC6A0F")) + 
  theme_minimal() +
  theme(plot.title = element_text(face = "bold"))

ggsave("PCA_non-embryonic.png", width = 10, height = 8)



# Tissue + cell type
pca_no_embryo <- plotPCA(vd_no_emb, intgroup = c("Tissue", "Cell.type"), returnData = TRUE)
perc_no_emb <- round(100 * attr(pca_no_embryo, "percentVar"))

p3_2 <- ggplot(pca_no_embryo, aes(PC1, PC2, color = Tissue, shape = Cell.type)) +
  geom_point(size = 3, stroke = 1) +
  labs(title = "PCA non-embryonic tissues",
       x = paste0("PC1 Var: ", perc_no_emb[1], " % variance"),
       y = paste0("PC2 Var: ", perc_no_emb[2], " % variance")) +
  scale_color_manual(values = c("#CBC118", "#3011F7", "#CC6A0F")) +
  scale_shape_manual(values = c(2, 0, 1, 17, 7, 8, 3, 15, 16)) +
  theme_minimal()

ggsave("PCA_non-embryonic_celltype.png", width = 15, height = 8)
```



**Combine plots**
```{r}
combined_plot <- (p1_2) / (p2 | p3) + 
  plot_layout(heights = c(1, 0.8))
ggsave("Combined_PCA_plots_tissue_celltype.png", plot = combined_plot, width = 12, height = 10, dpi = 300)

dev_tis_vs_cell_tis <- (p1_2 / p1_cell) +
    plot_annotation(title = "Initial PCAs",
                  theme = theme(plot.title = element_text(size = 12, face = "bold")),
                  tag_levels = "A")
ggsave("Combined_dev_tis-vs-cell_tis.png", plot = dev_tis_vs_cell_tis, width = 12, height = 10, dpi = 300)
```



## 4. Pre-filtering
```{r pre_filtering}
replicates <- table(metadata_removed$Tissue)
minimum_replicates <- min(replicates)
minimum_replicates

#### PRE-FILTERING
smallestGroupSize <- minimum_replicates
keep_group <- rowSums(counts(dds_105) > 10) >= smallestGroupSize
dds_105_filtered <- dds_105[keep_group,]
dds_105_filtered
```



## 4. Normalized counts to calculate CV and Gini Index
```{r}
dds_105_filtered <- estimateSizeFactors(dds_105_filtered)
expr_matrix_105 <- counts(dds_105_filtered, normalized = TRUE)
saveRDS(expr_matrix_106, file = "expr_matrix_106_normalized.rds")
```



## 5. Calculate the coefficient of variation (CV)
```{r calculate_cv}
# Mean expression per gene
expr_matrix_105 <- readRDS("expr_matrix_105_normalized.rds")
gene_means_105 <- rowMeans(expr_matrix_105)

# Dispersion of the data relative to the mean
gene_stdev_105 <- apply(expr_matrix_105, 1, sd) 

# Formula to calculate CV
cv_values_105 <- gene_stdev_105/gene_means_105 
```


```{r density_plot_cv_selection, warning=FALSE}
# Dataframe for ggplot
df_cv_values_105 <- data.frame(Gene = names(cv_values_105), 
                           CV = as.numeric(cv_values_105))

# Take the 2nd percentile (lowest 2%)
cv_threshold_105 <- quantile(df_cv_values_105$CV, probs = 0.02, na.rm = TRUE) 

# Select candidate housekeeping genes
genes_low_CV_105 <- cv_values_106[cv_values_105 < cv_threshold_105] # CV values below the threshold
genes_low_CV_105_names <- names(genes_low_CV_105) # Extract their names
length(genes_low_CV_105) # 519 genes


# Compute the density of the CV values
density_cv_105 <- density(df_cv_values_105$CV) # by dong this I obtain the minimum/maximum value, 1st/3rd quartile, mean, and media for x and y

# Value of 'x' in which 'y' is maximum
max_cv_105 <- density_cv_105$x[which.max(density_cv_105$y)] 



# DENSITY PLOT
ggplot(df_cv_values_105, aes(x = CV)) +
  geom_density(fill = "#009E73", alpha=0.2) +
  geom_vline(xintercept = max_cv_105, color = "#009E73", linetype = "dashed", size = 0.5) +
  geom_text(aes(x = max_cv_105, y = 0.7), label = round(max_cv_105, 3), color = "#009E73", hjust = -0.3, vjust = 1) +
  geom_vline(xintercept = cv_threshold_105, color = "#D55E00", linetype = "dashed", size = 0.5) +
  geom_text(aes(x = cv_threshold_105, y = 0.7), label = round(cv_threshold_105, 3), color = "#D55E00", hjust = 1, vjust = 0) +
  labs(title = "Density plot of CV values", x = "Coefficient of Variation", y = "Density") +
  theme_minimal()

ggsave("Density_plot_CV.png", width = 10, height = 8)
```



## 6. Calculate the Gini Index
Measures how stable is the expression of a gene across different samples. Uses the Lorenz Curve, which compares the cumulative distribution of gene expression across samples. Genes with low Gini Index are considered more stable.
```{r calculate_gini_index}
# Create the Gini Index function
gini_index <- function(expression_values){
  if (sum(expression_values) == 0) return (NA)
  
  expression_values <- sort(expression_values)
  n <- length(expression_values)
  
  cum_sum <- cumsum(expression_values)
  
  numerator <- sum((2*(1:n) - n - 1) * expression_values)
  denominator <- n * sum(expression_values)
  
  gini <- numerator/denominator
  
  return(gini)
}

gini_scores_105 <- apply(expr_matrix_105, 1, gini_index) 
```



```{r density_plot_gini}
gini_results_105 <- data.frame(Gene = names(gini_scores_105), 
                           Gini_Index = gini_scores_105)

gini_threshold_105 <- quantile(gini_results_105$Gini_Index, probs = 0.02, na.rm = TRUE) # 2%
genes_low_gini_105 <- gini_results_105[gini_results_105$Gini_Index < gini_threshold_105, ]
dim(genes_low_gini_105) # 519 x 2
genes_low_gini_105_names <- genes_low_gini_105$Gene


# compute the density of the gini values
density_gini_105 <- density(gini_results_105$Gini_Index) 

# value of x in which y is maximum
m_gini_105 <- density_gini_105$x[which.max(density_gini_105$y)] 


## DENSITY PLOT ##
ggplot(gini_results_105, aes(x = Gini_Index)) +
  geom_density(fill = "#009E73", alpha=0.2) +
  theme_minimal() +
  geom_vline(xintercept = m_gini_105, color = "#009E73", linetype = "dashed", size = 0.5) +
  geom_text(aes(x = m_gini_105, y = 1.5), label = round(m_gini_105, 3), color = "#009E73", hjust = -0.3, vjust = -0.5) +
  geom_vline(xintercept = gini_threshold_105, color = "#D55E00", linetype = "dashed", size = 0.5) +
  geom_text(aes(x = gini_threshold_105, y = 1.5), label = round(gini_threshold_105, 3), color = "#D55E00", hjust = 1.3, vjust = 0) +
  labs(title = "Density plot of Gini values", x = "Gini Index Values", y = "Density")

ggsave("Density_plot_Gini.png", width = 10, height = 8)
```



**Lorenz Curve**
```{r lorenz_curve}
ginis_sort_105 <- sort(gini_results_105$Gini_Index)
length(ginis_sort_105)

low_ginis_105 <- ginis_sort_105[1:3] # the first three
high_ginis_105 <- ginis_sort_105[25900:25902] # the last three

low_ginis_105_name <- subset(gini_results_105, Gini_Index %in% low_ginis_105)
high_ginis_105_name <- subset(gini_results_105, Gini_Index %in% high_ginis_105)

genes_low_ginis_105 <- low_ginis_105_name$Gene
genes_high_ginis_105 <- high_ginis_105_name$Gene

png("Lorenz_curves_genes.png", width = 1200, height = 800)
par(mfrow = c(2, 3), # 2 rows, 3 columns
    oma = c(5, 4, 5, 4), # (bottom, left, top, right)
    mgp = c(4, 1, 0), # (label, tick, line)
    mar = c(7, 8, 6, 5))  # (bottom, left, top, right) (between plots of the same row)

# Calculate and plot the Lorenz curve for the first 3 genes with the smallest value
top <- for (i in genes_low_ginis_105) {
  gene_expression_105 <- expr_matrix_105[i,]  # Obtain the expression of the i gene
  lorenz_curve_105 <- Lc(gene_expression_105)  # Calculate the Lorenz curve
  plot(lorenz_curve_105,
   	main = paste("Gene", i),
   	xlab = "Genes",
   	ylab = "Expression",
   	col = "#0072B2", lwd = 3, 
   	cex.main = 2,
   	cex.lab = 2,
   	cex.axis = 2,
   	family = "serif")
  abline(0, 1, col = "black", lty = 2)  # Add the diagonal (perfect equality)
}

# Calculate and plot the Lorenz curve for the last 3 genes with the highest value
down <- for (i in genes_high_ginis_105) {
  gene_expression_105 <- expr_matrix_105[i,]  # Obtain the expression of the i gene
  lorenz_curve_105 <- Lc(gene_expression_105)  # Calculate the Lorenz curve
  plot(lorenz_curve_105,
   	main = paste("Gene", i),
   	xlab = "Genes", # Accumulative proportion of genes
   	ylab = "Expression", # Accumulative proportion of expression
   	col = "#0072B2", lwd = 3,
   	cex.main = 2,
   	cex.lab = 2,
   	cex.axis = 2,
   	family = "serif")
  abline(0, 1, col = "black", lty = 2)  # Add the diagonal (perfect equality)
}

mtext("Lorenz curves for genes with top 3 low and high Gini Indices", outer = TRUE, cex = 2, line = 2, family = "serif") # cex is the size of the text
dev.off()

```



## 7. Select candidate HKGs and check if literature HKGs are in the candidates
```{r compare_lit_HKGs_candidates}
### LITERATURE HKGs ###
# B2M: ENSG00000166710
# ACTB: ENSG00000075624 
# GAPDH: ENSG00000111640
# HPRT1: ENSG00000165704
# HMBS: ENSG00000256269
# PPIA: ENSG00000196262
# SDHA: ENSG00000073578
# YWHAZ: ENSG00000164924
# TBP: ENSG00000112592
# RPLP0: ENSG00000089157
# UBC: ENSG00000150991
# GUSB: ENSG00000169919
# PGK1: ENSG00000102144
# RPL13A: ENSG00000142541


hkg_lit <- c("ENSG00000166710", "ENSG00000075624", "ENSG00000111640", "ENSG00000165704", "ENSG00000256269", "ENSG00000196262", "ENSG00000073578", "ENSG00000164924", "ENSG00000112592", "ENSG00000089157", "ENSG00000150991",  "ENSG00000169919", "ENSG00000102144", "ENSG00000142541")

# Check if the typical HKGs in literature are among the 519 selected by CV
intersect(genes_low_CV_105_names, hkg_lit) 

# Check if the typical HKGs in literature are among the 519 selected by Gini Index
intersect(genes_low_gini_105_names, hkg_lit) 

```
ENSG00000112592 appears in the CV list.
ENSG00000112592 and ENSG00000196262 appear in the Gini Index list.



**Extract CV and Gini of the literature HKGs**
```{r}
# Extract the CV values of the literature HKGs
cv_literature_genes_105 <- cv_values_105[hkg_lit] 
cv_literature_genes_105

# Extract the Gini index values of the literature HKGs
gini_literature_genes_105 <- gini_results_105[hkg_lit, ]
gini_literature_genes_105
```


**Extract the expression of the literature HKGs**
```{r}
df_exp_hkg <- data.frame(Gen = hkg_lit,
                         Media = sapply(hkg_lit, function(g) mean(as.numeric(counts_105[g, ]))),
                         Min = sapply(hkg_lit, function(g) min(counts_105[g, ])),
                         Max = sapply(hkg_lit, function(g) max(counts_105[g, ])))
```



**Extract the housekeeping genes selected in common: CV and Gini Index**
```{r select_common_HKGs}
# Common genes selected by both metrics
common_CV_Gini_105 <- intersect(genes_low_CV_105_names, genes_low_gini_105_names)
length(common_CV_Gini_105) # 433
saveRDS(common_CV_Gini_105, file = "final_candidate_HKGs_105.rds")

# Extract the expression of those candidate genes
housekeeping_data_105 <- expr_matrix_105[common_CV_Gini_105, ]

# Check if literature housekeeping genes in my candidates
intersect(common_CV_Gini_105, hkg_lit) 
```
ENSG00000112592 is the unique common gene between the literature list and mine.

**Extract the expression of my HKGs**
```{r}
df_exp_my_hkg <- data.frame(Gen = common_CV_Gini_105,
                         Media = sapply(common_CV_Gini_105, function(g) mean(as.numeric(counts_105[g, ]))),
                         Min = sapply(common_CV_Gini_105, function(g) min(counts_105[g, ])),
                         Max = sapply(common_CV_Gini_105, function(g) max(counts_105[g, ])))
```

```{r}
# Selected HKGs
expr_values <- rowMeans(counts_105[common_CV_Gini_105, ])
df <- data.frame(Expression = expr_values, Group = "Selected HKGs")

exp_selected_HKGs <- ggplot(df, aes(x = Group, y = Expression)) +
  geom_violin(fill = "#85eed0", alpha = 0.2) +
  geom_boxplot(width = 0.1, color = "black", outlier.size = 0.5) +
  scale_y_log10(limits = c(1, 1.4e6)) +
  labs(x = "", y = "Expression media") +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 11, color = "black"))

# Literature HKGs
expr_values_lit1 <- rowMeans(counts_105[hkg_lit, ])
df_lit1 <- data.frame(Expression = expr_values_lit1, Group = "Literature HKGs")

exp_lit_HKGs <- ggplot(df_lit1, aes(x = Group, y = Expression)) +
  geom_violin(fill = "#eee885", alpha = 0.2) +
  geom_boxplot(width = 0.1, color = "black", outlier.size = 0.5) +
  scale_y_log10(limits = c(1, 1.4e6)) +
  labs(x = "", y = "Expression media") +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 11, color = "black"))

# All genes
df_all <- data.frame(Expression = rowMeans(counts_105), Group = "All genes")

exp_all <- ggplot(df_all, aes(x = Group, y = Expression)) +
  geom_violin(fill = "#85afee", alpha = 0.2) +
  geom_boxplot(width = 0.1, color = "black", outlier.size = 0.5) +
  scale_y_log10(limits = c(1, 1.4e6)) +
  labs(title = "Expression", x = "", y = "Expression media") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold"),
        axis.text.x = element_text(size = 11, color = "black"))


# Combine plot: expression of all genes, the literature ones, and mine
combined_dexpressions <- (exp_all / exp_lit_HKGs / exp_selected_HKGs) +
    plot_annotation(title = "Distribution of expression", 
                  theme = theme(plot.title = element_text(size = 16, face = "bold")),
                  tag_levels = "A")
ggsave("Combined_expressions.png", width = 12, height = 14)
```



**Order the literature HKGs**
```{r rank_lit_HKGs}
df_hkg <- data.frame(Gene = hkg_lit,
                     CV = as.numeric(cv_hkg_lit),
                     Gini = as.numeric(gini_literature_genes_105$Gini_Index))

# Calculate the ranking for each metric
df_hkg$rank_CV <- rank(df_hkg$CV, ties.method = "min")
df_hkg$rank_Gini <- rank(df_hkg$Gini, ties.method = "min")

# Media of the ranking
df_hkg$mean_rank <- rowMeans(df_hkg[, c("rank_CV", "rank_Gini")])

# Select top 14 (all)
all_hkg <- df_hkg[order(df_hkg$mean_rank), ][1:14, ]
all_hkg
```


**Density plots CV values**
```{r density_plots_CV_values}
### All genes ###
# Calculate statistics
mean_all <- mean(df_cv_values_105$CV)
median_all <- median(df_cv_values_105$CV)
mode_all <- density(df_cv_values_105$CV)$x[which.max(density(df_cv_values_105$CV)$y)]

# Prepare the dataframe
df_lines_all <- data.frame(
  Statistic = c("Mean", "Median", "Mode"),
  Value = c(mean_all, median_all, mode_all)
)

df_lines_all$label <- round(df_lines_all$Value, 2)
dens_all <- density(df_cv_values_105$CV)
max_y_all <- max(dens_all$y)
df_lines_all$y <- max_y_all * 0.9

# Create the density plot for all the genes
p1_all <- ggplot(df_cv_values_105, aes(x = CV)) +
  geom_density(fill = "#85afee", alpha = 0.2) +
  geom_vline(data = df_lines_all,
             aes(xintercept = Value, color = Statistic),
             linewidth = 0.8) +
  scale_color_manual(values = c("Mean" = "#97201e",
                                "Median" = "#1e5897",
                                "Mode" = "#1e9782")) +
  labs(title = "All genes",
       x = "CV",
       y = "Density",
       color = "Statistic") +
  scale_x_continuous(breaks = round(df_lines_all$Value, 2)) +
  theme_minimal()
ggsave("Density_plot_CV_all_genes.png", width = 10, height = 8)



### Selected candidate HKGs ###
# Calculate the CV of the selected candidate genes
cv_candidates <- cv_values_105[common_CV_Gini_105]
df_cv_candidates <- data.frame(CV = cv_candidates)

# Calculate statistics
mean_cand <- mean(cv_candidates)
median_cand <- median(cv_candidates)
mode_cand <- density(cv_candidates)$x[which.max(density(cv_candidates)$y)]

df_lines_cand <- data.frame(Statistic = c("Mean", "Median", "Mode"),
                           Value = c(mean_cand, median_cand, mode_cand))

df_lines_cand$label <- round(df_lines_cand$Value, 2)
dens_cand <- density(df_cv_candidates$CV)
max_y_cand <- max(dens_cand$y)
df_lines_cand$y <- max_y_cand * 0.9


# Plot
p2_candidate <- ggplot(df_cv_candidates, aes(x = CV)) +
  geom_density(fill = "#85eed0", alpha = 0.2) +
  geom_vline(data = df_lines_cand,
             aes(xintercept = Value, color = Statistic),
             linewidth = 0.8, show.legend = TRUE) +
  scale_color_manual(values = c("Mean" = "#97201e",
                                "Median" = "#1e5897",
                                "Mode" = "#1e9782")) +
  labs(title = "HKGs selected",
       x = "CV",
       y = "Density",
       color = "Statistic") +
  scale_x_continuous(breaks = round(df_lines_cand$Value, 2)) +
  theme_minimal()
ggsave("Density_plot_CV_candidate_genes.png", width = 10, height = 8)



### Literature HKGs ###
# Calculate the CVs for the literature HKGs
cv_hkg_lit <- cv_values_105[hkg_lit]
df_cv_hkg_lit <- data.frame(CV = cv_hkg_lit)

# Calculate statistics
mean_lit <- mean(cv_hkg_lit)
median_lit <- median(cv_hkg_lit)
mode_lit <- density(cv_hkg_lit)$x[which.max(density(cv_hkg_lit)$y)]

df_lines_lit <- data.frame(
  Statistic = c("Mean", "Median", "Mode"),
  Value = c(mean_lit, median_lit, mode_lit)
)

df_lines_lit$label <- round(df_lines_lit$Value, 2)
dens_lit <- density(cv_hkg_lit)
max_y_lit <- max(dens_lit$y)
df_lines_lit$y <- max_y_lit * 0.9

# Density plot
p3_lit <- ggplot(df_cv_hkg_lit, aes(x = CV)) +
  geom_density(fill = "#eee885", alpha = 0.2) +
  geom_vline(data = df_lines_lit,
             aes(xintercept = Value, color = Statistic),
             linewidth = 0.8) +
  scale_color_manual(values = c("Mean" = "#97201e",
                                "Median" = "#1e5897",
                                "Mode" = "#1e9782")) +
  labs(title = "HKGs literature", x = "CV", y = "Density") +
  scale_x_continuous(breaks = round(df_lines_lit$Value, 2)) +
  theme_minimal()
ggsave("Density_plot_CV_literature_genes.png", width = 10, height = 8)


# Combine the three density plots in a unique one
combined_plot_CV <- p1_all / p2_candidate / p3_lit +
  plot_annotation(title = "Coefficient of Variation Distributions", 
                  theme = theme(plot.title = element_text(size = 16, face = "bold")),
                  tag_levels = "A")
ggsave("Combined_density_plots_CV.png", plot = combined_plot_CV, width = 12, height = 10)
```


```{r}
### Create a dataframe containing the CV values of the different groups
# Groups: all genes, literature HKGs, selected candidate HKGs
df_combined_cv <- data.frame(CV = c(cv_values_105, cv_literature_genes_105, cv_candidates), Group = factor(c(rep("All genes", length(cv_values_105)), rep("HKGs literature", length(cv_literature_genes_105)), rep("HKGs selected", length(cv_candidates)))))

# Density plot to compare the CV distribution across the groups
ggplot(df_combined_cv, aes(x=CV, fill=Group, color=Group)) +
  geom_density(alpha=0.4) +
  theme_minimal() +
  labs(x = "Coefficient of  Variation (CV)", y = "Density",
       title = "CV comparison distributions") +
  scale_fill_manual(values=c("#85afee", "#85eed0", "#eee885")) +
  scale_color_manual(values=c("#85afee", "#85eed0", "#eee885"))
ggsave("Density_plot_all_in_one.png", width = 10, height = 8)


# Boxplot to visualize and compare the CV dispersion across the groups
b_cv <- ggplot(df_combined_cv, aes(x=Group, y=CV, fill=Group)) +
  geom_boxplot(alpha=0.6, outlier.size=0.7) +
  annotate("text", x = 1.3, y = 2.4, label = "n = 25902", size = 2.9) +
  annotate("text", x = 2.2, y = 1.1, label = "n = 14", size = 2.9) +
  annotate("text", x = 3.2, y = 0.6, label = "n = 433", size = 2.9) +
  theme_minimal() +
  labs(title = "CV",
       x = "", y = "Coefficient of  Variation (CV)") +
  scale_fill_manual(values=c("#85afee", "#85eed0", "#eee885")) +
  theme(legend.position = "none")
```



**Violin plots Gini Indices**
```{r}
range(gini_results_105$Gini_Index)

# All genes
g_violin_all <- ggplot(gini_results_105, aes(x = " ", y = Gini_Index)) +
  geom_violin(fill = "#85afee", alpha = 0.2) +
  geom_boxplot(width = 0.1, color = "black", outlier.size = 0.5) +
  coord_cartesian(ylim = c(0, 1)) +
  labs(title = "Gini Index", x = "Gini Index all genes", y = "Density") +
  theme_minimal() + 
  theme(plot.title = element_text(face = "bold"))


# Candidate HKGs
g_violin_cand <- ggplot(gini_candidates, aes(x = " ", y = Gini_Index)) +
  geom_violin(fill = "#85eed0", alpha = 0.2) +
  geom_boxplot(width = 0.1, color = "black", outlier.size = 0.5) +
  coord_cartesian(ylim = c(0, 1)) +
  labs(x = "Gini Index selected HKGs", y = "Density") +
  theme_minimal()


# Literature HKGs
g_violin_lit <- ggplot(gini_hkg, aes(x = " ", y = Gini_Index)) +
  geom_violin(fill = "#eee885", alpha = 0.2) +
  geom_boxplot(width = 0.1, color = "black", outlier.size = 0.5) +
  coord_cartesian(ylim = c(0, 1)) +
  labs(x = "Gini Index literature HKGs", y = "Density") +
  theme_minimal()
```


**Violin plots CV**
```{r}
# All genes
cv_values_all <- data.frame(CV = cv_values_105)

cv_violin_all <- ggplot(cv_values_all, aes(x = " ", y = CV)) +
  geom_violin(fill = "#85afee", alpha = 0.2) +
  geom_boxplot(width = 0.1, color = "black", outlier.size = 0.5) +
  labs(title = "CV", x = "CV all genes", y = "Density") +
  coord_cartesian(ylim = c(0, 3)) +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold"),
        axis.text.x = element_text(size = 11))


# Candidate HKGs
cv_values_common_cand <- data.frame(CV = cv_values_common)

cv_violin_cand <- ggplot(cv_values_common_cand, aes(x = " ", y = CV)) +
  geom_violin(fill = "#85eed0", alpha = 0.2) +
  geom_boxplot(width = 0.1, color = "black", outlier.size = 0.5) +
  labs(x = "CV selected HKGs", y = "Density") +
  coord_cartesian(ylim = c(0, 3)) +
  theme_minimal()


# Literature HKGs
cv_hkg_lit_lit <- data.frame(CV = cv_hkg_lit)

cv_violin_lit <- ggplot(cv_hkg_lit_lit, aes(x = " ", y = CV)) +
  geom_violin(fill = "#eee885", alpha = 0.2) +
  geom_boxplot(width = 0.1, color = "black", outlier.size = 0.5) +
  labs(x = "CV literature HKGs", y = "Density") +
  coord_cartesian(ylim = c(0, 3)) + 
  theme_minimal()
```


**Combine expression, CV, Gini Index for all three groups**
```{r plot_combined_violin_exp_cv_gini}
# Combine expression plots
c_exp <- exp_all + exp_lit_HKGs + exp_selected_HKGs

# Combine CV plots
c_cv <- cv_violin_all + cv_violin_lit + cv_violin_cand

# Combine Gini Index plots
c_gini <- g_violin_all + g_violin_lit + g_violin_cand

# Create the combined figure
c <- (c_exp / plot_spacer() / c_cv / plot_spacer() / c_gini) +   
  plot_layout(heights = c(1, 0.1, 1, 0.1, 1)) +
  plot_annotation(title = "Comparison of expression level, Coefficient of Variation, and Gini Index", theme = theme(plot.title = element_text(size = 16, face = "bold", hjust = 0.5)), tag_levels = "A")

ggsave("Combined_exp_cv_gini.png", width = 12, height = 15)
```



```{r combined_gini_boxplot}
## Create the dataframe containing the Gini values of the different groups
df_combined_gini <- data.frame(Gini = c(gini_results_105$Gini_Index, gini_literature_genes_105$Gini_Index, gini_candidates$Gini_Index),
                                    Group = factor(c(rep("All genes", length(gini_results_105$Gini_Index)),
                                                     rep("HKGs literature", length(gini_literature_genes_105$Gini_Index)),
                                                     rep("HKGs selected", length(gini_candidates$Gini_Index)))))


b_gini <- ggplot(df_combined_gini, aes(x = Group, y = Gini, fill=Group)) +
  geom_boxplot(alpha=0.6, outlier.size=0.7) +
  annotate("text", x = 1.3, y = 0.8, label = "n = 25902", size = 2.9) +
  annotate("text", x = 2.2, y = 0.4, label = "n = 14", size = 2.9) +
  annotate("text", x = 3.2, y = 0.2, label = "n = 433", size = 2.9) +
  theme_minimal() +
  labs(title = "Gini Index",
       x = "", y = "Gini Index") +
  scale_fill_manual(values=c("#85afee", "#85eed0", "#eee885")) +
  theme(legend.position = "none")


combined_boxplots_cv_gini <- b_cv | b_gini
ggsave("Combined_boxplots_cv_gini.png", width = 10, height = 6)
```



**Venn diagram CV and Gini Index**
```{r}
png("VennDiagram_candidate_HKGs.png")
par(mar = c(5, 5, 5, 5)) # inferior, izquierda, superior, derecha
draw.pairwise.venn(
  area1 = length(genes_low_CV_105_names),
  area2 = length(genes_low_gini_105_names),
  cross.area = length(intersect(genes_low_CV_105_names, genes_low_gini_105_names)),
  category = c("CV", "Gini"),
  fill = c("#1AFF1A", "#4B0092"),
  cex = 2,
  cat.cex = 2.2
)
dev.off()
```



## 8. GO Enrichment Analysis of the candidate HKGs
We performed ORA Enrichment Analysis using the final set of housekeeping genes selected by both methods (CV and Gini Index). This analysis identifies overrepresented Gene Ontology (GO) terms associated with biological processes (BP).

By default:
  pvalueCutoff = 0.05
  qvalueCutoff = 0.2
  
```{r EA_candidate_hkg}
housekeeping_genes <- common_CV_Gini_105
all_expressed_genes <- rownames(expr_matrix_105)

go_enrich_hkg_BP <- enrichGO(gene = housekeeping_genes,
                      OrgDb = org.Hs.eg.db,
                      keyType = 'ENSEMBL',
                      ont = "BP",
                      universe = all_expressed_genes,
                      readable = TRUE, # convert gene IDs to gene symbols
                      pvalueCutoff = 0.05, # the one by default
                      qvalueCutoff = 0.2) # the one by default

View(data.frame(go_enrich_hkg_BP))
```

**View the top 20**
```{r}
# Convertir a data frame
go_enrich_hkg_BP_df <- as.data.frame(go_enrich_hkg_BP)

# Ordenar por p-valor ajustado (de menor a mayor)
go_enrich_hkg_BP_df_sorted <- go_enrich_hkg_BP_df[order(go_enrich_hkg_BP_df$p.adjust), ]

# Extraer las top 20
top20_go_terms <- head(go_enrich_hkg_BP_df_sorted, 20)

# Ver el resultado
View(top20_go_terms)

```


```{r visualization_EA_candidate_HKG}
# For quick insights: BARPLOT
#png("Barplot_top_BP.png", width = 1600, height = 1200, res = 300)
png("Barplot_HKG.png", height = 20, width = 18, units = "cm", res = 300)
barplot(go_enrich_hkg_BP, showCategory = 12) + 
  theme(axis.text.y = element_text(angle = 0, hjust = 1, size = 10)) +
  ggtitle("Top Biological Processes")
dev.off()

# For gene term connections: CNET PLOT
png("Cnetplot_HKG.png", width = 1600, height = 1200, res = 300)
cnetplot(go_enrich_hkg_BP, showCategory = 12) + 
  theme(axis.text.y = element_text(angle = 0, hjust = 1, size = 10)) +
  ggtitle("Cnetplot of top Biological Processes") 
dev.off()
```






## 9. HSC vs Dendritic progenitors
```{r}
# 1. Extract the metadata
cells_of_interest <- c("hematopoietic stem cells (CD34+lin-CD38-CD45RA-CD90+)",
                       "common dendritic progenitors (CD34+lin-CD38+CD45RA+CD10-CD123hi)")
metadata_hsc_dendritic <- metadata_removed[metadata_removed$Cell.type %in% cells_of_interest, ]

# 2. Extract the counts
selected_hsc_dendritic <- rownames(metadata_hsc_dendritic) # Extract the names of the selected samples
counts_hsc_dendritic <- counts_105[, selected_hsc_dendritic] # Filter counts_106_removed by those columns
dim(counts_hsc_dendritic) # 63103 x 16

```

**Construct the DESeqDataSet() object**
```{r}
dds_hsc_dendritic <- DESeqDataSetFromMatrix(countData = counts_hsc_dendritic,
                                  colData = metadata_hsc_dendritic,
                                  design = ~ Cell.type) 
```

**Pre-filtering**
```{r}
replicates_hsc_dendritic <- table(metadata_hsc_dendritic$Tissue)
min_rep_HSC_dendritic <- min(replicates_hsc_dendritic)
min_rep_HSC_dendritic

#### PRE-FILTERING
smallestGroupSize <- min_rep_HSC_dendritic
keep_group <- rowSums(counts(dds_hsc_dendritic) > 10) >= smallestGroupSize
dds_hsc_dendritic_filtered <- dds_hsc_dendritic[keep_group,]
dds_hsc_dendritic_filtered
```

**Differential Expression Analysis**
```{r run_deseq}
# Execute DESeq to estimate dispersion and adjust the model
dds_hsc_dendritic <- DESeq(dds_hsc_dendritic_filtered)

res_hsc_dendritic <- resultsNames(dds_hsc_dendritic) 
res_hsc_dendritic
# There are 2 coefficients
```

**Comparison of interest**
```{r}
res_HSC_vs_dendritic <- results(dds_hsc_dendritic,
                           contrast = c("Cell.type",
                                        "hematopoietic stem cells (CD34+lin-CD38-CD45RA-CD90+)",
                                        "common dendritic progenitors (CD34+lin-CD38+CD45RA+CD10-CD123hi)"),
                           independentFiltering = FALSE)
head(res_HSC_vs_dendritic, 3)
summary(res_HSC_vs_dendritic)
```

**RUV correction**
```{r ruvg_preparation}
# Define control genes (the 430 HKGs selected)
controls_105 <- readRDS("final_candidate_HKGs_105.rds")

# Raw counts of the filtered genes
counts_filtered_hsc_dendritic <- counts(dds_hsc_dendritic)

# Metadata 
pheno_hsc_dendritic <- as.data.frame(colData(dds_hsc_dendritic))

# Check colnames and rownames are the same
rownames(pheno_hsc_dendritic) == colnames(counts_filtered_hsc_dendritic)

# Create the object from the count matrix
seq_hsc_dendritic <- newSeqExpressionSet(counts = as.matrix(counts_filtered_hsc_dendritic),
                           phenoData = pheno_hsc_dendritic)

# Remove unwanted variation with the control genes
# K is the number of unwanted variation factors to remove
ruv_k1_hsc_dendritic <- RUVg(seq_hsc_dendritic, controls_105, k = 1)
ruv_k2_hsc_dendritic <- RUVg(seq_hsc_dendritic, controls_105, k = 2)

# Extract Ws and add them to the metadata
metadata_w1_hsc_dendritic <- pData(ruv_k1_hsc_dendritic)
metadata_w1_hsc_dendritic$W_1 <- pData(ruv_k1_hsc_dendritic)$W_1

###### K2 just for check
metadata_w2_hsc_dendritic <- pData(ruv_k2_hsc_dendritic)
metadata_w2_hsc_dendritic$W_1 <- pData(ruv_k2_hsc_dendritic)$W_1
metadata_w2_hsc_dendritic$W_2 <- pData(ruv_k2_hsc_dendritic)$W_2
```


**Comparison between DESeq2 and DESeq2 with RUV**
```{r}
png("plotRLE_hsc_dendritic.png")
par(mfrow = c(1,3), las = 2, mar = c(7, 4, 4, 2))
  
plotRLE(seq_hsc_dendritic, outline=FALSE, ylim=c(-3, 3), main = "Before RUV") # Before RUV
plotRLE(ruv_k1_hsc_dendritic, outline=FALSE, ylim=c(-3, 3), main = "RUV with k=1") # RUV with k=1
plotRLE(ruv_k2_hsc_dendritic, outline=FALSE, ylim=c(-3, 3), main = "RUV with k=2") # RUV with k=2
dev.off()
```
Choose K=1



**PCA after RUV**
```{r}
# Check rownames and colnames are the same
rownames(metadata_w1_hsc_dendritic) == colnames(ruv_k1_hsc_dendritic)

dds_hsc_dendritic_ruv1 <- DESeqDataSetFromMatrix(countData = counts(ruv_k1_hsc_dendritic),
                                  colData   = metadata_w1_hsc_dendritic,
                                  design    = ~ W_1 + Cell.type)
```


**Differential Expression Analysis with RUV (k=1)**
```{r}
# Adjust the model with DESeq
dds_ruv1_hsc_dendritic <- DESeq(dds_hsc_dendritic_ruv1)
res_hsc_dendritic_ruv1 <- results(dds_ruv1_hsc_dendritic)

# Extract the results
res_HSC_vs_dendritic_ruv1 <- results(dds_ruv1_hsc_dendritic,
                          contrast = c("Cell.type",
                                        "hematopoietic stem cells (CD34+lin-CD38-CD45RA-CD90+)",
                                        "common dendritic progenitors (CD34+lin-CD38+CD45RA+CD10-CD123hi)"),
                           independentFiltering = FALSE)
head(res_HSC_vs_dendritic_ruv1, 3)
summary(res_HSC_vs_dendritic_ruv1)

```

**Selecting significant genes**
```{r}
### BEFORE RUV
DE_genes <- res_HSC_vs_dendritic[which(res_HSC_vs_dendritic$padj < 0.05), ]
up <- DE_genes[DE_genes$log2FoldChange > 1, ]
down <- DE_genes[DE_genes$log2FoldChange < -1, ]
```

```{r}
### AFTER RUV
DE_genes_ruv <- res_HSC_vs_dendritic_ruv1[which(res_HSC_vs_dendritic_ruv1$padj < 0.05), ]
up_ruv <- DE_genes_ruv[DE_genes_ruv$log2FoldChange > 1, ]
down_ruv <- DE_genes_ruv[DE_genes_ruv$log2FoldChange < -1, ]
```


**Enrichment analysis**
```{r}
## UP before RUV
up_genes <- rownames(up)

go_up <- enrichGO(gene = up_genes,
                  OrgDb = org.Hs.eg.db,
                  keyType = "ENSEMBL",
                  ont = "BP",
                  universe = rownames(res_HSC_vs_dendritic),
                  readable = TRUE,
                  pvalueCutoff = 0.05,
                  qvalueCutoff = 0.2)

View(data.frame(go_up)) # 133 enriched terms

## DOWN before RUV
down_genes <- rownames(down)

go_down <- enrichGO(gene = down_genes,
                  OrgDb = org.Hs.eg.db,
                  keyType = "ENSEMBL",
                  ont = "BP",
                  universe = rownames(res_HSC_vs_dendritic),
                  readable = TRUE,
                  pvalueCutoff = 0.05,
                  qvalueCutoff = 0.2)

View(data.frame(go_down)) # 640 enriched terms
```

```{r}
## UP after RUV
up_genes_ruv <- rownames(up_ruv)

go_up_ruv <- enrichGO(gene = up_genes_ruv,
                  OrgDb = org.Hs.eg.db,
                  keyType = "ENSEMBL",
                  ont = "BP",
                  universe = rownames(res_HSC_vs_dendritic_ruv1),
                  readable = TRUE,
                  pvalueCutoff = 0.05,
                  qvalueCutoff = 0.2)

View(data.frame(go_up_ruv)) # 135 enriched terms

## DOWN after RUV
down_genes_ruv <- rownames(down_ruv)

go_down_ruv <- enrichGO(gene = down_genes_ruv,
                  OrgDb = org.Hs.eg.db,
                  keyType = "ENSEMBL",
                  ont = "BP",
                  universe = rownames(res_HSC_vs_dendritic_ruv1),
                  readable = TRUE,
                  pvalueCutoff = 0.05,
                  qvalueCutoff = 0.2)

View(data.frame(go_down_ruv)) # 650 enriched terms
```

```{r}
### UP before RUV
# For quick insights: BARPLOT
png("Barplot_up.png", height = 20, width = 18, units = "cm", res = 300)
up_before_ruv <- barplot(go_up, showCategory = 12) + theme(axis.text.y = element_text(angle = 0, hjust = 1, size = 11), plot.title = element_text(size = 4)) + ggtitle("UP (before RUV)")
dev.off()

### DOWN before RUV
# For quick insights: BARPLOT
png("Barplot_down.png", height = 20, width = 18, units = "cm", res = 300)
down_before_ruv <- barplot(go_down, showCategory = 12) + theme(axis.text.y = element_text(angle = 0, hjust = 1, size = 11), plot.title = element_text(size = 4)) + ggtitle("DOWN (before RUV)")
dev.off()


### UP after RUV
# For quick insights: BARPLOT
png("Barplot_up_ruv.png", height = 20, width = 18, units = "cm", res = 300)
up_after_ruv <- barplot(go_up_ruv, showCategory = 12) + theme(axis.text.y = element_text(angle = 0, hjust = 1, size = 11), plot.title = element_text(size = 4)) + ggtitle("UP (after RUV)")
dev.off()

### DOWN after RUV
# For quick insights: BARPLOT
png("Barplot_down_ruv.png", height = 20, width = 18, units = "cm", res = 300)
down_after_ruv <- barplot(go_down_ruv, showCategory = 12) + theme(axis.text.y = element_text(angle = 0, hjust = 1, size = 11), plot.title = element_text(size = 4)) + ggtitle("DOWN (after RUV)")
dev.off()


# Combine all barplots (up, down, and ruv, no ruv) into a single figure
combined_barplots_hsc_dendritic <- (up_before_ruv | up_after_ruv) / (down_before_ruv | down_after_ruv) +
  plot_annotation(title = "Top enriched biological processes – HSC vs Dendritic progenitors", tag_levels = "A") &
  theme(plot.title = element_text(size = 20, face = "bold", hjust = 0.5),
        plot.margin = margin(t = 20, r = 10, b = 10, l = 10))  # top, right, bottom, left
ggsave("Combined_barplots_hsc_dendritic.png", width = 17, height = 16, dpi = 450)
```


**Check unique and common terms**
```{r checking_common_unique_terms}
## UP
common_up <- intersect(go_up$Description, go_up_ruv$Description) # 87 common terms for up
only_before_up <- setdiff(go_up$Description, go_up_ruv$Description) # 46 unique terms before applying correction in the up-regulated
only_after_up <- setdiff(go_up_ruv$Description, go_up$Description) # 48 unique terms after applying correction in the down-regulated 

## DOWN
common_down <- intersect(go_down$Description, go_down_ruv$Description) # 575 common terms for down
only_before_down <- setdiff(go_down$Description, go_down_ruv$Description) # 65 unique terms before applying correction in the down-regulated
only_after_down <- setdiff(go_down_ruv$Description, go_down$Description) # 75 unique terms after applying correction in the down-regulated
```

**Extract the full object of the previously extracted terms**
```{r}
go_up_only_before <- go_up
go_up_only_before@result <- filter(go_up@result, Description %in% only_before_up)

go_up_only_after <- go_up_ruv
go_up_only_after@result <- filter(go_up_ruv@result, Description %in% only_after_up)

go_down_only_before <- go_down
go_down_only_before@result <- filter(go_down@result, Description %in% only_before_down)

go_down_only_after <- go_down_ruv
go_down_only_after@result <- filter(go_down_ruv@result, Description %in% only_after_down)
```


**Visualization of the top enriched terms**
```{r visualization_EA_up_down}
### UP before RUV
# For quick insights: BARPLOT
png("Barplot_up_unique.png", height = 20, width = 18, units = "cm", res = 300)
barplot(go_up_only_before, showCategory = 12) + theme(axis.text.y = element_text(angle = 0, hjust = 1, size = 10)) + ggtitle("Top UP unique biological processes (before RUV)")
dev.off()


### DOWN before RUV
# For quick insights: BARPLOT
png("Barplot_down_unique.png", height = 20, width = 18, units = "cm", res = 300)
barplot(go_down_only_before, showCategory = 12) + theme(axis.text.y = element_text(angle = 0, hjust = 1, size = 10)) + ggtitle("Top DOWN unique biological processes (before RUV)")
dev.off()


### UP after RUV
# For quick insights: BARPLOT
png("Barplot_up_unique_ruv.png", height = 20, width = 18, units = "cm", res = 300)
barplot(go_up_only_after, showCategory = 12) + theme(axis.text.y = element_text(angle = 0, hjust = 1, size = 10)) + ggtitle("Top UP unique biological processes (after RUV)")
dev.off()


### DOWN after RUV
# For quick insights: BARPLOT
png("Barplot_down_unique_ruv.png", height = 20, width = 18, units = "cm", res = 300)
barplot(go_down_only_after, showCategory = 12) + theme(axis.text.y = element_text(angle = 0, hjust = 1, size = 10)) + ggtitle("Top DOWN unique biological processes (after RUV)")
dev.off()

```






## 10. Hematopoietic progenitor vs Endothelium
```{r}
# 1. Extract the metadata
cells_of_interest_2 <- c("endothelium (CD34+CD90+CD43-CD45-)",
                       "hematopoietic progenitors (CD34+CD90- (CD43+CD45+))")
metadata_hemato_endo <- metadata_removed[metadata_removed$Cell.type %in% cells_of_interest_2, ]

# 2. Extract the counts
selected_hemato_endo <- rownames(metadata_hemato_endo) # Extract the names of the selected samples
counts_hemato_endo <- counts_106_removed[, selected_hemato_endo] # Filter counts_106_removed by those columns
dim(counts_hemato_endo) # 63103 x 26

```

**Construct the DESeqDataSet() object**
```{r}
dds_hemato_endo <- DESeqDataSetFromMatrix(countData = counts_hemato_endo,
                                  colData = metadata_hemato_endo,
                                  design = ~ Cell.type) 
```

**Pre-filtering**
```{r}
replicates_hemato_endo <- table(metadata_hemato_endo$Tissue)
min_rep_hemato_endo <- min(replicates_hemato_endo)
min_rep_hemato_endo

#### PRE-FILTERING
smallestGroupSize <- min_rep_hemato_endo
keep_group <- rowSums(counts(dds_hemato_endo) > 10) >= smallestGroupSize
dds_hemato_endo_filtered <- dds_hemato_endo[keep_group,]
dds_hemato_endo_filtered
```

**Differential Expression Analysis**
```{r run_deseq}
# Execute DESeq to estimate dispersion and adjust the model
dds_hemato_endo <- DESeq(dds_hemato_endo_filtered)

res_hemato_endo <- resultsNames(dds_hemato_endo) 
res_hemato_endo
```

**Comparison of interest**
```{r}
res_hemato_vs_endo <- results(dds_hemato_endo,
                           contrast = c("Cell.type", "hematopoietic progenitors (CD34+CD90- (CD43+CD45+))", "endothelium (CD34+CD90+CD43-CD45-)"),
                           independentFiltering = FALSE)
head(res_hemato_vs_endo, 3)
summary(res_hemato_vs_endo)
```

**Remove Unwanted Variation**
```{r remove_unwanted_variation}
# Define control genes (the 430 HKGs selected)
controls_106 <- common_CV_Gini_105

# Raw counts of the filtered genes
counts_filtered_hemato_endo <- counts(dds_hemato_endo)

# Metadata 
pheno_hemato_endo <- as.data.frame(colData(dds_hemato_endo))

# Check colnames and rownames are the same
rownames(pheno_hemato_endo) == colnames(counts_filtered_hemato_endo)

# Create the object from the count matrix
seq_hemato_endo <- newSeqExpressionSet(counts = as.matrix(counts_filtered_hemato_endo),
                           phenoData = pheno_hemato_endo)

# Remove unwanted variation with the control genes
# K is the number of unwanted variation factors to remove
ruv_k1_hemato_endo <- RUVg(seq_hemato_endo, controls_105, k = 1)
ruv_k2_hemato_endo <- RUVg(seq_hemato_endo, controls_105, k = 2) # just to check

# Extract Ws and add them to the metadata
metadata_w1_hemato_endo <- pData(ruv_k1_hemato_endo)
metadata_w1_hemato_endo$W_1 <- pData(ruv_k1_hemato_endo)$W_1

###### K2 just for check
metadata_w2_hemato_endo <- pData(ruv_k2_hemato_endo)
metadata_w2_hemato_endo$W_1 <- pData(ruv_k2_hemato_endo)$W_1
metadata_w2_hemato_endo$W_2 <- pData(ruv_k2_hemato_endo)$W_2
```

**Comparison between DESeq2 and DESeq2 with RUV**
```{r}
png("plotRLE_hemato_endo.png")
par(mfrow = c(1,3), las = 2, mar = c(7, 4, 4, 2))
  
plotRLE(seq_hemato_endo, outline=FALSE, ylim=c(-3, 3), main = "Before RUV") # Before RUV
plotRLE(ruv_k1_hemato_endo, outline=FALSE, ylim=c(-3, 3), main = "RUV with k=1") # RUV with k=1
plotRLE(ruv_k2_hemato_endo, outline=FALSE, ylim=c(-3, 3), main = "RUV with k=2") # RUV with k=2
dev.off()
```
Choose k= 1


**PCA after RUV**
```{r}
# Check rownames and colnames are the same
rownames(metadata_w1_hemato_endo) == colnames(ruv_k1_hemato_endo)

dds_hemato_endo_ruv1 <- DESeqDataSetFromMatrix(countData = counts(ruv_k1_hemato_endo),
                                  colData   = metadata_w1_hemato_endo,
                                  design    = ~ W_1 + Cell.type)
```


**Differential Expression Analysis with RUV (k=1)**
```{r}
# Adjust the model with DESeq
dds_ruv1_hemato_endo <- DESeq(dds_hemato_endo_ruv1)
res_hemato_vs_endo_ruv1 <- results(dds_ruv1_hemato_endo)

# Extract the results
res_hemato_vs_endo_ruv1 <- results(dds_ruv1_hemato_endo,
                           contrast = c("Cell.type",
                                        "hematopoietic progenitors (CD34+CD90- (CD43+CD45+))",
                                        "endothelium (CD34+CD90+CD43-CD45-)"),
                           independentFiltering = FALSE)
head(res_hemato_vs_endo_ruv1, 3)
summary(res_hemato_vs_endo_ruv1)
```

**Selecting significant genes**
```{r}
### BEFORE RUV
DE_genes_hemato_endo <- res_hemato_vs_endo[which(res_hemato_vs_endo$padj < 0.05), ]
up_hemato_endo <- DE_genes_hemato_endo[DE_genes_hemato_endo$log2FoldChange > 1, ]
down_hemato_endo <- DE_genes_hemato_endo[DE_genes_hemato_endo$log2FoldChange < -1, ]
```

```{r}
### AFTER RUV
DE_genes_hemato_endo_ruv <- res_hemato_vs_endo_ruv1[which(res_hemato_vs_endo_ruv1$padj < 0.05), ]
up_hemato_endo_ruv <- DE_genes_hemato_endo_ruv[DE_genes_hemato_endo_ruv$log2FoldChange > 1, ]
down_hemato_endo_ruv <- DE_genes_hemato_endo_ruv[DE_genes_hemato_endo_ruv$log2FoldChange < -1, ]
```


**Enrichment analysis**
```{r}
## UP before RUV
up_genes_hemato_endo <- rownames(up_hemato_endo)

go_up_hemato_endo <- enrichGO(gene = up_genes_hemato_endo,
                  OrgDb = org.Hs.eg.db,
                  keyType = "ENSEMBL",
                  ont = "BP",
                  universe = rownames(res_hemato_vs_endo),
                  readable = TRUE,
                  pvalueCutoff = 0.05,
                  qvalueCutoff = 0.2) 

View(data.frame(go_up_hemato_endo)) # 1118 enriched terms

## DOWN before RUV
down_genes_hemato_endo <- rownames(down_hemato_endo)

go_down_hemato_endo <- enrichGO(gene = down_genes_hemato_endo,
                  OrgDb = org.Hs.eg.db,
                  keyType = "ENSEMBL",
                  ont = "BP",
                  universe = rownames(res_hemato_vs_endo),
                  readable = TRUE,
                  pvalueCutoff = 0.05,
                  qvalueCutoff = 0.2) 

View(data.frame(go_down_hemato_endo)) # 1328 enriched terms
```

```{r}
## UP after RUV
up_genes_ruv_hemato_endo <- rownames(up_hemato_endo_ruv)

go_up_ruv_hemato_endo <- enrichGO(gene = up_genes_ruv_hemato_endo,
                  OrgDb = org.Hs.eg.db,
                  keyType = "ENSEMBL",
                  ont = "BP",
                  universe = rownames(res_hemato_vs_endo_ruv1),
                  readable = TRUE,
                  pvalueCutoff = 0.05,
                  qvalueCutoff = 0.2) 

View(data.frame(go_up_ruv_hemato_endo)) # 1130 enriched terms

## DOWN after RUV
down_genes_ruv_hemato_endo <- rownames(down_hemato_endo_ruv)

go_down_ruv_hemato_endo <- enrichGO(gene = down_genes_ruv_hemato_endo,
                  OrgDb = org.Hs.eg.db,
                  keyType = "ENSEMBL",
                  ont = "BP",
                  universe = rownames(res_hemato_vs_endo_ruv1),
                  readable = TRUE,
                  pvalueCutoff = 0.05,
                  qvalueCutoff = 0.2) 

View(data.frame(go_down_ruv_hemato_endo)) # 1263 enriched terms
```


**Visualization of the top enriched terms**
```{r}
png("Barplot_up.png")
up_before_ruv_hemato_endo <- barplot(go_up_hemato_endo, showCategory = 12) + theme(axis.text.y = element_text(angle = 0, hjust = 1, size = 11), plot.title = element_text(size = 4)) + ggtitle("UP (before RUV)")
dev.off()

png("Barplot_down.png")
down_before_ruv_hemato_endo <- barplot(go_down_hemato_endo, showCategory = 12) + theme(axis.text.y = element_text(angle = 0, hjust = 1, size = 11), plot.title = element_text(size = 4)) + ggtitle("DOWN (before RUV)")
dev.off()

png("Barplot_up_ruv.png")
up_after_ruv_hemato_endo <- barplot(go_up_ruv_hemato_endo, showCategory = 12) + theme(axis.text.y = element_text(angle = 0, hjust = 1, size = 11), plot.title = element_text(size = 4)) + ggtitle("UP (after RUV)")
dev.off()

png("Barplot_down_ruv.png")
down_after_ruv_hemato_endo <- barplot(go_down_ruv_hemato_endo, showCategory = 12) + theme(axis.text.y = element_text(angle = 0, hjust = 1, size = 11), plot.title = element_text(size = 4)) + ggtitle("DOWN (after RUV)")
dev.off()


# Combine the barplots into a single figure
combined_barplots_hemato_endo <- (up_before_ruv_hemato_endo | up_after_ruv_hemato_endo) / (down_before_ruv_hemato_endo | down_after_ruv_hemato_endo) +
  plot_annotation(title = "Top enriched biological processes – Hematopoietic progenitors vs Endothelium", tag_levels = "A") &
  theme(plot.title = element_text(size = 20, face = "bold", hjust = 0.5),
        plot.margin = margin(t = 20, r = 10, b = 10, l = 10))  # top, right, bottom, left
ggsave("Combined_barplots_hemato_endo.png", width = 17, height = 16, dpi = 450)

```

**Checking unique and common enriched terms**
```{r check_unique_common_enriched_terms}
## UP
common_up_hemato_endo <- intersect(go_up_hemato_endo$Description, go_up_ruv_hemato_endo$Description) # 1052 common enriched terms up-regulated
only_before_up_hemato_endo <- setdiff(go_up_hemato_endo$Description, go_up_ruv_hemato_endo$Description) # 66 unique terms before applying correction in the up-regulated
only_after_up_hemato_endo <- setdiff(go_up_ruv_hemato_endo$Description, go_up_hemato_endo$Description) # 78 unique terms after applying correction in the up-regulated

## DOWN
common_down_hemato_endo <- intersect(go_down_hemato_endo$Description, go_down_ruv_hemato_endo$Description) # 1193 common enriched terms down-regulated
only_before_down_hemato_endo <- setdiff(go_down_hemato_endo$Description, go_down_ruv_hemato_endo$Description) # 135 unique terms before applying correction in the down-regulated
only_after_down_hemato_endo <- setdiff(go_down_ruv_hemato_endo$Description, go_down_hemato_endo$Description) # 70 unique terms after applying correction in the down-regulated
```

**Extract the full object of the previously extracted terms**
```{r}
go_up_only_before_hemato_endo <- go_up_hemato_endo
go_up_only_before_hemato_endo@result <- filter(go_up_hemato_endo@result, Description %in% only_before_up_hemato_endo)

go_up_only_after_hemato_endo <- go_up_ruv_hemato_endo
go_up_only_after_hemato_endo@result <- filter(go_up_ruv_hemato_endo@result, Description %in% only_after_up_hemato_endo)

go_down_only_before_hemato_endo <- go_down_hemato_endo
go_down_only_before_hemato_endo@result <- filter(go_down_hemato_endo@result, Description %in% only_before_down_hemato_endo)

go_down_only_after_hemato_endo <- go_down_ruv_hemato_endo
go_down_only_after_hemato_endo@result <- filter(go_down_ruv_hemato_endo@result, Description %in% only_after_down_hemato_endo)
```

**Visualizations of the terms**
```{r go_barplots_up_down_unique}
### UP before RUV
# For quick insights: BARPLOT
png("Barplot_up_unique.png")
barplot(go_up_only_before_hemato_endo, showCategory = 12) + theme(axis.text.y = element_text(angle = 0, hjust = 1, size = 7)) + ggtitle("Top UP unique biological processes (before RUV)")
dev.off()


### DOWN before RUV
# For quick insights: BARPLOT
png("Barplot_down_unique.png")
barplot(go_down_only_before_hemato_endo, showCategory = 12) + theme(axis.text.y = element_text(angle = 0, hjust = 1, size = 7)) + ggtitle("Top DOWN unique biological processes (before RUV)")
dev.off()


### UP after RUV
# For quick insights: BARPLOT
png("Barplot_up_ruv_unique.png")
barplot(go_up_only_after_hemato_endo, showCategory = 12) + theme(axis.text.y = element_text(angle = 0, hjust = 1, size = 7)) + ggtitle("Top UP unique biological processes (after RUV)")
dev.off()


### DOWN after RUV
# For quick insights: BARPLOT
png("Barplot_down_ruv_unique.png")
barplot(go_down_only_after_hemato_endo, showCategory = 12) + theme(axis.text.y = element_text(angle = 0, hjust = 1, size = 7)) + ggtitle("Top DOWN unique biological processes (after RUV)")
dev.off()
```



## 11. Differentially expressed genes before and after RUV correction

**Comparison 1: Haematopoietic progenitors vs Endothelium**
```{r}
up_counts <- 2830
up_counts_ruv <- 3018

down_counts <- 6243
down_counts_ruv <- 5907


df <- data.frame(
  Method = c("No RUV", "No RUV", "RUV", "RUV"),
  Direction = c("Up", "Down", "Up", "Down"),
  Count = c(up_counts, -down_counts, up_counts_ruv, -down_counts_ruv)
)

up_down_hemato_endo <- ggplot(df, aes(x = Method, y = Count, fill = Direction)) +
  geom_col(width = 0.5) +
  geom_hline(yintercept = 0, color = "black") +
  scale_fill_manual(values = c("Up" = "#5DADE2", "Down" = "#E74C3C")) +
  geom_text(aes(label = abs(Count), vjust = ifelse(Count > 0, -0.3, 1.3)), size = 4) +
  ylab("Number of DE genes") +
  xlab("") +
  theme_minimal() +
  ggtitle("Up and Down Regulated Genes (Hematopoietic progenitors vs Endothelium)") +
  theme(axis.text.x = element_text(size = 12), plot.title = element_text(face = "bold"))

```


**Comparison 2: HSC vs Dendritic progenitors**
```{r}
up_counts <- 4039
up_counts_ruv <- 6294

down_counts <- 2415
down_counts_ruv <- 2679

df <- data.frame(
  Method = c("No RUV", "No RUV", "RUV", "RUV"),
  Direction = c("Up", "Down", "Up", "Down"),
  Count = c(up_counts, -down_counts, up_counts_ruv, -down_counts_ruv)
)

up_down_hsc_dend <- ggplot(df, aes(x = Method, y = Count, fill = Direction)) +
  geom_col(width = 0.5) +
  geom_hline(yintercept = 0, color = "black") +
  scale_fill_manual(values = c("Up" = "#5DADE2", "Down" = "#E74C3C")) +
  geom_text(aes(label = abs(Count), vjust = ifelse(Count > 0, -0.3, 1.3)), size = 4) +
  ylab("Number of DE genes") +
  xlab("") +
  theme_minimal() +
  ggtitle("Up and Down Regulated Genes (HSC vs Dendritic progenitors)") +
  theme(axis.text.x = element_text(size = 12), plot.title = element_text(face = "bold"))
```

```{r}
combined_up_down <- up_down_hemato_endo / up_down_hsc_dend
ggsave("Combined_hist_up_down.png", width = 12, height = 14)
```



**Hemato vs Endo**
```{r}
# Put together up and down for each condition
relevant_genes_no_ruv <- union(rownames(up_hemato_endo), rownames(down_hemato_endo))
relevant_genes_ruv <- union(rownames(up_hemato_endo_ruv), rownames(down_hemato_endo_ruv))

venn1 <- venn.diagram(x = list(No_RUV = relevant_genes_no_ruv,
                               RUV = relevant_genes_ruv),
  #filename = "venn_hemato_vs_endo.png",
  filename = NULL,
  fill = c("#5DADE2", "#58D68D"),
  alpha = 0.5,
  cex = 1,
  cat.cex = 1,
  main = "Relevant DE genes Hematopoietic progenitors vs Endothelium\n(|log2FC| > 1 & padj < 0.05)")
```


**HSC vs Dendritic**
```{r}
# Put together up and down for each condition
relevant_genes_no_ruv_hsc_Den <- union(rownames(up), rownames(down))
relevant_genes_ruv_hsc_Den <- union(rownames(up_ruv), rownames(down_ruv))

venn2 <- venn.diagram(x = list(No_RUV = relevant_genes_no_ruv_hsc_Den,
                               RUV = relevant_genes_ruv_hsc_Den),
  #filename = "venn_HSC_vs_dendritic.png",
  filename = NULL,
  fill = c("#5DADE2", "#58D68D"),
  alpha = 0.5, # color transparency
  cex = 1, # size of the numbers
  cat.cex = 1, # size of the text
  main = "Relevant DE genes HSC vs Dendritic progenitors\n(|log2FC| > 1 & padj < 0.05)")
```

```{r}
blank_space <- textGrob(" ", gp = gpar(fontsize = 20))

png("combined_venn_diagrams.png", width = 2000, height = 3000, res = 300)
grid.arrange(
  grobTree(venn1),
  blank_space,
  grobTree(venn2),
  ncol = 1,
  heights = c(1, 0.1, 1)
)
dev.off()
```


```{r}
combined_up_down <- up_down_hemato_endo / up_down_hsc_dend

# Convert the venn diagrams to plots with cowplot
venn_plot1 <- ggdraw(grobTree(venn1))
venn_plot2 <- ggdraw(grobTree(venn2))

venn_combined <- venn_plot1 / plot_spacer() / venn_plot2 + plot_layout(heights = c(1, 0.1, 1)) # 0.1 is the space between the venn diagrams

final_plot <- combined_up_down | venn_combined

ggsave("final_histograms_venns.png", final_plot, width = 16, height = 12, dpi = 300)
```


12. Enriched terms
**Hematopoietic progenitors vs Endothelium**
```{r}
# go_up_hemato_endo
# go_up_ruv_hemato_endo
up_terms <- 1118
up_terms_ruv <- 1130

# go_down_hemato_endo
# go_down_ruv_hemato_endo
down_terms <- 1328
down_terms_ruv <- 1263


df <- data.frame(
  Method = c("No RUV", "No RUV", "RUV", "RUV"),
  Direction = c("Up", "Down", "Up", "Down"),
  Terms_enriched = c(up_terms, -down_terms, up_terms_ruv, -down_terms_ruv)
)

up_down_hemato_endo <- ggplot(df, aes(x = Method, y = Terms_enriched, fill = Direction)) +
  geom_col(width = 0.5) +
  geom_hline(yintercept = 0, color = "black") +
  scale_fill_manual(values = c("Up" = "#5DADE2", "Down" = "#E74C3C")) +
  geom_text(aes(label = abs(Terms_enriched), vjust = ifelse(Terms_enriched > 0, -0.3, 1.3)), size = 4) +
  ylab("Number of Enriched Terms") +
  xlab("") +
  theme_minimal() +
  ggtitle("Up and Down BP (Hematopoietic progenitors vs Endothelium)") +
  theme(axis.text.x = element_text(size = 12), plot.title = element_text(face = "bold"))

```

**HSC vs Dendritic progenitors**
```{r}
# go_up
# go_up_ruv
up_terms <- 133
up_terms_ruv <- 135

# go_down
# go_down_ruv
down_terms <- 640
down_terms_ruv <- 650

df <- data.frame(
  Method = c("No RUV", "No RUV", "RUV", "RUV"),
  Direction = c("Up", "Down", "Up", "Down"),
  Terms_enriched = c(up_terms, -down_terms, up_terms_ruv, -down_terms_ruv)
)

up_down_hsc_dend <- ggplot(df, aes(x = Method, y = Terms_enriched, fill = Direction)) +
  geom_col(width = 0.5) +
  geom_hline(yintercept = 0, color = "black") +
  scale_fill_manual(values = c("Up" = "#5DADE2", "Down" = "#E74C3C")) +
  geom_text(aes(label = abs(Terms_enriched), vjust = ifelse(Terms_enriched > 0, -0.3, 1.3)), size = 4) +
  ylab("Number of Enriched Terms") +
  xlab("") +
  theme_minimal() +
  ggtitle("Up and Down BP (HSC vs Dendritic progenitors)") +
  theme(axis.text.x = element_text(size = 12), plot.title = element_text(face = "bold"))
```

```{r}
combined_up_down <- up_down_hemato_endo / up_down_hsc_dend
ggsave("Combined_hist_up_down_terms.png", width = 12, height = 14)
```

**Hemato vs Endo**
```{r}
# Put together up and down for each condition
up_before_hemato_endo <- go_up_hemato_endo@result$ID[go_up_hemato_endo@result$p.adjust < 0.05]
up_after_hemato_endo  <- go_up_ruv_hemato_endo@result$ID[go_up_ruv_hemato_endo@result$p.adjust < 0.05]

down_before_hemato_endo <- go_down_hemato_endo@result$ID[go_down_hemato_endo@result$p.adjust < 0.05]
down_after_hemato_endo  <- go_down_ruv_hemato_endo@result$ID[go_down_ruv_hemato_endo@result$p.adjust < 0.05]

total_before_hemato_endo <- union(up_before_hemato_endo, down_before_hemato_endo)
total_after_hemato_endo  <- union(up_after_hemato_endo, down_after_hemato_endo)


venn1_terms <- venn.diagram(x = list(No_RUV = total_before_hemato_endo,
                               RUV = total_after_hemato_endo),
  #filename = "venn_hemato_vs_endo.png",
  filename = NULL,
  fill = c("#5DADE2", "#58D68D"),
  alpha = 0.5,
  cex = 1,
  cat.cex = 1,
  main = "Enriched terms Hematopoietic progenitors vs Endothelium\n(|log2FC| > 1 & padj < 0.05)")
```


**HSC vs Dendritic**
```{r}
# Put together up and down for each condition
up_before <- go_up@result$ID[go_up@result$p.adjust < 0.05]
up_after  <- go_up_ruv@result$ID[go_up_ruv@result$p.adjust < 0.05]

down_before <- go_down@result$ID[go_down@result$p.adjust < 0.05]
down_after  <- go_down_ruv@result$ID[go_down_ruv@result$p.adjust < 0.05]

total_before <- union(up_before, down_before)
total_after  <- union(up_after, down_after)


venn2_terms <- venn.diagram(x = list(No_RUV = total_before,
                               RUV = total_after),
  #filename = "venn_HSC_vs_dendritic.png",
  filename = NULL,
  fill = c("#5DADE2", "#58D68D"),
  alpha = 0.5, # color transparency
  cex = 1, # size of the numbers
  cat.cex = 1, # size of the text
  main = "Enriched terms HSC vs Dendritic progenitors\n(|log2FC| > 1 & padj < 0.05)")
```

```{r}
blank_space <- textGrob(" ", gp = gpar(fontsize = 20))

png("combined_venn_diagrams_terms.png", width = 2000, height = 3000, res = 300)
grid.arrange(
  grobTree(venn1_terms),
  blank_space,
  grobTree(venn2_terms),
  ncol = 1,
  heights = c(1, 0.1, 1)
)
dev.off()
```

```{r}
combined_up_down_terms <- up_down_hemato_endo / up_down_hsc_dend

# Convert the venn diagrams to plots with cowplot
venn_plot1_terms <- ggdraw(grobTree(venn1_terms))
venn_plot2_terms <- ggdraw(grobTree(venn2_terms))

venn_combined_terms <- venn_plot1_terms / plot_spacer() / venn_plot2_terms + plot_layout(heights = c(1, 0.1, 1)) # 0.1 is the space between the venn diagrams

final_plot_terms <- combined_up_down_terms | venn_combined_terms

ggsave("final_histograms_venns_terms.png", final_plot_terms, width = 16, height = 12, dpi = 300)
```




12. Density distributions of CV and Gini Index highlighting literature HKGs
**Density curve of CV values with lit HKGs marked**
```{r}
df_cv_all <- cv_values_all

# Create a dataframe with the CV values of the literature HKGs
df_cv_lit <- data.frame(CV = cv_hkg_lit_lit, Gene = rownames(cv_hkg_lit_lit))

# Convert the IDs of the genes to a gene name
ensembl_ids <- df_cv_lit$Gene
gene_symbols <- mapIds(org.Hs.eg.db,
                       keys = ensembl_ids,
                       column = "SYMBOL",
                       keytype = "ENSEMBL",
                       multiVals = "first")
df_cv_lit$Gene <- gene_symbols

# Filter the dataframe to only have information about TBP
df_tbp <- df_cv_lit[df_cv_lit$Gene == "TBP", ]

# Density plot of all genes adding lines where the literature HKGs are located
density_cv <- ggplot(df_cv_all, aes(x = CV)) +
  geom_density(fill = "#b5d3e7", alpha = 0.5, color = "black") +
  geom_vline(data = df_cv_lit, aes(xintercept = CV, color = Gene == "TBP"), 
             linetype = "dashed", size = 0.9) +
  scale_color_manual(values = c("FALSE" = "#e74c3c", "TRUE" = "#1abc9c"), 
                     labels = c("Other HKGs", "TBP")) +
  geom_text(data = df_tbp, aes(x = CV, y = max(density(df_cv_all$CV)$y) + 0.02, label = Gene),
            angle = 45, hjust = 1, size = 3.2, color = "#1abc9c") +
  geom_vline(xintercept = cv_threshold_106, color = "black", linetype = "dashed", size = 1) +
  geom_text(aes(x = cv_threshold_106, y = max(density(df_cv_all$CV)$y) - 0.02), label = round(cv_threshold_106, 3), color = "black", hjust = 1, vjust = 0, angle = 45, size = 3.2) +
  labs(title = "Distribution of CV values across all genes",
       x = "Coefficient of Variation (CV)",
       y = "Density",
       color = NULL) +
  theme_minimal() +
  theme(plot.title = element_text(size = 14, face = "bold"),
        plot.subtitle = element_text(size = 10),
        legend.position = "top") +
  guides(color = "none")

```



**Density curve of Gini Index values with lit HKGs marked**
```{r}
## Exactly the same as before but now for the Gini Index instead of the CV

df_gini_all <- gini_results_106
df_gini_lit <- gini_literature_genes_106

ensembl_ids <- df_gini_lit$Gene
gene_symbols <- mapIds(org.Hs.eg.db,
                       keys = ensembl_ids,
                       column = "SYMBOL",
                       keytype = "ENSEMBL",
                       multiVals = "first")
df_gini_lit$Gene <- gene_symbols

genes_interest <- c("TBP", "PPIA")
df_labels <- df_gini_lit[df_gini_lit$Gene %in% genes_interest, ]

max_y_gini <- max(density(df_gini_all$Gini_Index)$y)

density_gini <- ggplot(df_gini_all, aes(x = Gini_Index)) +
  geom_density(fill = "#b5d3e7", alpha = 0.5, color = "black") +
  geom_vline(data = df_gini_lit, aes(xintercept = Gini_Index), 
             linetype = "dashed", color = "#e74c3c", size = 0.9) +
  geom_vline(data = df_labels, aes(xintercept = Gini_Index), linetype = "dashed", size = 0.9, color = "#1abc9c") +
  geom_text(data = df_labels, aes(x = Gini_Index, y = max_y_gini + 0.02, label = Gene),
            angle = 45, hjust = 1, size = 3.2, color = "#1abc9c") +
  geom_vline(xintercept = gini_threshold_106, color = "black", linetype = "dashed", size = 1) +
  geom_text(aes(x = gini_threshold_106, y = max_y_gini - 0.07), 
            label = round(gini_threshold_106, 3), color = "black", 
            hjust = 1, vjust = 0, angle = 45, size = 3.2) +
  labs(title = "Distribution of Gini Index values across all genes",
       x = "Gini Index",
       y = "Density",
       color = NULL) +
  theme_minimal() +
  theme(plot.title = element_text(size = 14, face = "bold"),
        plot.subtitle = element_text(size = 10),
        legend.position = "top")

```


```{r}
combined_densities <- density_cv / density_gini
ggsave("Combined_densities.png", width = 14, height = 10)
```




